% Quant II Recitation
% Drew Dimmery <drewd@nyu.edu>
% January 31, 2014

# What is this?

- This presentation created using [knitr](http://yihui.name/knitr/) + [pandoc](http://johnmacfarlane.net/pandoc/) + [reveal.js](http://lab.hakim.se/reveal-js/)
- I have created a recitation repository on github: [https://github.com/ddimmery/Quant-II-Recitation](https://github.com/ddimmery/Quant-II-Recitation)
- This presentation is available at [https://github.com/ddimmery/Quant-II-Recitation/raw/master/2014-01-31/presentation.html](https://github.com/ddimmery/Quant-II-Recitation/blob/master/2014-01-31/presentation.html)

# Today's Topic

- Sharing is caring.
    - If you don't share your research (+ code and data) then what's the point of doing it?
- Today we will be discussing some necessary mechanics.
    - R and how to bend it to your will
    - Producing high quality tables / plots
    - If there's any time left, we can talk a little more about identification

# Sharing
- All homework will be submitted to me (preferably in hard copy)
- All replication code will be posted to a gist on github, and submitted to me via email.
    - [gist.github.com](https://gist.github.com/)
    - Code should have a header with relevant information (name, date created, date modified, input files, output files, etc)
    - **Code should be well commented**
    - If you'd prefer to submit homework as some sort of knitr document, that is also fine. Just submit the `.Rmd` file.
- All tables and plots should be of very high quality.
- Yes, this will take a non-trivial amount of time.

# Workflow
- Find an editor and learn everything about it:
    - **[vim](http://www.vim.org/)** (+ [Vim-R-plugin](http://www.vim.org/scripts/script.php?script_id=2628))
    - [emacs](http://www.gnu.org/software/emacs/) (+ [ESS](http://ess.r-project.org/) [Emacs Speaks Statistics])
    - [Notepad++](http://notepad-plus-plus.org/), [Kate](http://kate-editor.org/), [Sublime](http://www.sublimetext.com/), etc
    - **[Rstudio](http://www.rstudio.com/)**
    - [Eclipse](http://www.eclipse.org/) (+ [StatET](http://www.walware.de/goto/statet))
- Familiarize yourself with version control software
    - [git](http://git-scm.com/) ([github](https://github.com/) or [TortoiseGit](https://code.google.com/p/tortoisegit/))
    - or just [Dropbox](https://www.dropbox.com/)

# When things break
- Documentation - Ex: `?lm`
- [Google](http://google.com)
- CRAN (Reference manuals, vignettes, etc) - Ex: [http://cran.r-project.org/web/packages/AER/index.html](http://cran.r-project.org/web/packages/AER/index.html)
- JSS - Ex: [http://www.jstatsoft.org/v27/i02](http://www.jstatsoft.org/v27/i02)
- Stack Overflow - [http://stackoverflow.com/questions/tagged/r](http://stackoverflow.com/questions/tagged/r)
- Listservs - [http://www.r-project.org/mail.html](http://www.r-project.org/mail.html)

# Resources
- [The Art of R Programming](http://heather.cs.ucdavis.edu/~matloff/132/NSPpart.pdf) - N. Matloff
- [Modern Applied Statistics with S](http://www.stats.ox.ac.uk/pub/MASS4/) - W. Venables and B. Ripley
- [Advanced R Programming](http://adv-r.had.co.nz/) - forthcoming, H. Wickham
- [The R Inferno](http://www.burns-stat.com/documents/books/the-r-inferno/) - P. Burns
- [Rdataviz](https://github.com/pablobarbera/Rdataviz) - a talk by P. Barberá on ggplot2

# Reading R Documentation
- <a href="http://stat.ethz.ch/R-manual/R-patched/library/stats/html/lm.html" target="_blank">`?lm`</a>

# CRAN documentation
- <a href="http://cran.r-project.org/web/packages/AER/index.html" target="_blank">AER</a>

# JSS
- <a href="http://www.jstatsoft.org/v27/i02/">`plm`</a>

# Confusing Things
- At the prompt, > means you’re good to go, + means a parenthesis or bracket is open.
- Case sensitive
- Use / in path names. Not \\.
- R uses variables – there is no "sheet" here, like in Stata
- **R is a programming language**
- More on errors later!

# Using Third-party Code
- Relevant commands are: `install.packages` and `library`
- Find the appropriate packages and commands with Google and via searching in R:
```
?covariance
??covariance
install.packages("sandwich")
library("sandwich")
?vcovHC
```

# Data types
- Character - strings
- Double / Numeric - numbers
- Logical - true/false
- Factor - unordered categorical variables
- Objects - its complicated

# Character

. . .

```{r}
str <- "This is a string"
```

. . .

```{r}
paste("This","is","a","string", sep=" ")
```

. . .

```{r}
as.character(99)
class(str)
```

# Numeric

. . .

```{r}
num <- 99.867
class(num)
```
. . .

```{r}
round(num, digits=2)
```

. . .

```{r}
round(str, digits=2)
```

. . .

```{r}
pi
exp(1)
```

- `sin`, `exp`, `log`, `factorial`, `choose`, `BesselJ`, etc

# Logical

- The logical type allows us to make statements about truth

. . .

```{r}
2 == 4
class(2==4)
```

. . .

```{r}
str != num
```

. . .

```{r}
"34" == 34
```

- `==`, `!=`, `>`, `<`, `>=`, `<=`, `!`, `&`, `|`, `any`, `all`, etc

# Objects

- Many functions will return objects rather than a single datatype.

. . .

```{r}
X <- 1:100; Y <- rnorm(100,X)
out.lm <- lm(Y~X)
class(out.lm)
```

- Objects can have other data embedded inside them

. . .

```{r}
out.lm$rank
class(out.lm$rank)
```

# Data Structures

- There are other ways to hold data, though:

    - Vectors
    - Lists
    - Matrices
    - Dataframes

# Vectors

- Almost everything in R is a vector.

. . .

```{r}
as.vector(4)
4
```

. . .

- We can combine vectors with `c`:

. . .

```{r}
vec <- c("a","b","c")
vec
```

. . .

```{r}
c(2,3,vec)
```

# Vectors (cont.)

- Sometimes R does some weird stuff:

. . .

```{r}
c(1,2,3,4) + c(1,2)
```

- It "recycles" the shorter vector:

. . .

```{r}
c(1,2,3,4) + c(1,2,1,2)
```

. . .

```{r}
c(1,2,3,4) + c(1,2,3)
```

# More Vectors
- We can index vectors in several ways

. . .

```{r}
vec[1]
```

. . .

```{r}
names(vec) <- c("first","second","third")
vec
```

. . .

```{r}
vec["first"]
```

# Missingness

. . .

```{r}
vec[1] <- NA
vec
```

. . .

```{r}
is.na(vec)
```

. . .
```{r}
vec[!is.na(vec)] # vec[complete.cases(vec)]
```

# Lists

- Lists are similar to vectors, but they allow for arbitrary mixing of types and lengths.

. . .

```{r}
listie <- list(first = vec, second = num)
listie
```

. . .

```{r}
listie[[1]]
listie$first
```

# Matrices

- $$A = \begin{pmatrix}1 & 3\\ 2 & 4\end{pmatrix}$$
- $A_{ij}$
- $A_{1,2} = 3$
- $A_{1,\cdot} = (1,3)$

. . .

```{r}
A <- matrix(c(1,2,3,4),nrow=2,ncol=2)
A
A[1,2]
A[1,]
```

# Matrix Operations

- Its very easy to manipulate matrices:

. . .

```{r}
solve(A) #A^{-1}
```

. . . 

```{r}
10*A
```

. . .

```{r}
B<-diag(c(1,2))
B
```

. . .

```{r}
A%*%B
```

# More Matrix Ops.

. . .

```{r}
A%*%diag(3)
```

```{r}
t(A) # A'
```

. . .

```{r}
rbind(A,B)
```

```{r}
cbind(A,B)
```

. . .

```{r}
c(1,2,3)%x%c(1,1) # Kronecker Product
```

# Naming Things

. . .

```{r}
rownames(A)
```

. . .

```{r}
rownames(A)<-c("a","b")
colnames(A)<-c("c","d")
A
```

. . .

```{r}
A[,"d"]
```

- Matrices are vectors:

. . .

```{r}
A[3]
```

# Dataframes

- The workhorse

- Basically just a matrix that allows mixing of types.

. . .

```{r}
data(iris)
head(iris)
```

# Control Flow

- loops
- if/else
- functions
- useful stock functions to know

# Loops

- for loops - just a way to say "do this for each element of the index"
- "this" is defined in what follows the "for" expression

. . .

```{r}
for(i in 1:5) {
  cat(i*10," ")
}
```

. . .

```{r}
for(i in 1:length(vec)) { 
  cat(vec[i]," ")
}
```

. . .

```{r}
for(i in vec) { 
  cat(i," ")
}
```

# If/Else

. . .

```{r}
if(vec[2]=="b") print("Hello World!")
```

. . .

```{r}
if(vec[3]=="a") {
  print("Hello World!")
} else {
  print("!dlroW olleH")
}
```

# Vectorized If/Else

- Conditional execution on each element of a vector

. . .

```{r}
vec <- letters[1:3]
new <- vector(length=length(vec))
for(i in 1:length(vec)) {
  if(vec[i]=="b") {
    new[i] <- 13
  } else {
    new[i] <- 0
  }
}
new
```

. . .

```{r}
new <- ifelse(vec=="b",13,0)
new
```

# Functions

- $f : X \to Y$
- Functions in R are largely the same. ("Pure functions")

. . .

```{r}
add3 <- function(X) {
  return(X+3)
}
add3(2)
```

. . .

```{r}
makeGroups<-function(groups,members=1) {
  return((1:groups)%x%rep(1,members))
}
makeGroups(5)
makeGroups(5,2)
```

# Useful Functions

- Note: Most functions don’t do complete case analysis by default (usually option na.rm=TRUE)

- `print`, `cat`, `paste`, `with`, `length`, `sort`, `order`, `unique`, `rep`, `nrow`, `ncol`, `complete.cases`, `subset`, `merge`, `mean`, `sum`, `sd`, `var`, `lag`,`lm`, `model.matrix`,`coef`, `vcov`, `residuals`, `vcovHC` (from `sandwich`), `ivreg` (from `AER`), `countrycode` (from`countrycode`),`summary`, `pdf`, `plot`, Tools from `plm`, and many more.

# Distributional Functions

- `?Distributions`
- They have a consistent naming scheme.
- `rnorm`, `dnorm`, `qnorm`, `pnorm`
- `rdist` - generate random variable from dist
- `ddist` - density function of dist
- `qdist` - quantile function of dist
- `pdist` - distribution function of dist
- look at documentation for parameterization

. . .

```{r}
rnorm(16)
```

# Robust SEs
```{.r .numberLines}
robust.se <- function(model,cluster=1:length(model$residuals)) {
  require(sandwich)
  require(lmtest)
  M <- length(unique(cluster))
  N <- length(cluster)
  K <- model$rank
  dfc <- (M/(M - 1)) * ((N - 1)/(N - K))
  uj  <- apply(
               estfun(model),
               2,
               function(x) tapply(x, cluster, sum)
              )
  rcse.cov <- dfc * sandwich(model,meat = crossprod(uj)/N)
  rcse.se <- coeftest(model, rcse.cov)
  return(list(rcse.cov, rcse.se))
}
```

# Multiple Dispactch

- Sometimes functions will behave differently based on context:

. . .

```{r}
summary(vec)
```

. . .

```{r}
summary(c(1,2,3,4))
```

. . .

```{r}
summary(iris[,1:4])
```

- Why? `?summary` `?summary.lm`
  
# The \*apply family

- These functions allow one to *efficiently* perform a large number of actions on data.
- `apply` - performs actions on the rows or columns of a matrix/array (1 for rows, 2 for columns, 3 for ??)
- `sapply` - performs actions on every element of a vector
- `tapply` - performs actions on a vector by group
- `replicate` - performs the same action a given number of times

# apply

```{r}
A
apply(A,1,sum)
apply(A,2,mean)
```

# sapply

```{r}
vec
sapply(vec,function(x) paste0(x,".vec"))
```
- Can be accomplished more simply with:

. . .

```{r}
paste0(vec,".vec")
```

- Why?

- `replicate` is basically just `sapply(1:N,funct)` where `funct` never uses the index.

# tapply

```{r}
tapply(1:10, makeGroups(5,2), mean)
```

# Working With Data

- Input
- Output

# Input

. . .

```{r}
setwd("~/github/Quant II Recitation/2014-01-31/")
dir()
iris <- read.csv("iris.csv")
```

- If data is, for instance, a Stata .dta file, use `read.dta` from the `foreign` package.

- Useful options for reading data: `sep`, `na.strings`, `stringsAsFactors`

- For different formats, Google it.

# Simulate some Data

```{r graphics,fig.cap='', fig.width=4.5, fig.height=4.5 }
set.seed(1023) # Important for replication
X <- rnorm(1000,0,5)
Y <- sin(5*X)*exp(abs(X)) + rnorm(1000)
dat <- data.frame(X,Y)
plot(X,Y,xlim=c(0,5),ylim=c(-50,50))
```

# Regression Output

```{r}
dat.lm<-lm(Y~X,data=dat)
dat.lm
```

# Regression Output

```{r}
summary(dat.lm)
```

# Pretty Output
- How do we get $\LaTeX$ output?
- The `xtable` package:

. . .

```{r}
require(xtable)
xtable(dat.lm)
```

# xtable
- `xtable` works on any sort of matrix

. . .

```{r}
xtable(A)
```

# xtable
- This is what `xtable` does with the `lm` object:

. . .

```{r}
class(summary(dat.lm)$coefficients)
xtable(summary(dat.lm)$coefficients)
```

- Note that this is the same as the output from `xtable(dat.lm)`

# Pretty it up

- Now let's make some changes to what `xtable` spits out:

. . .

```{r}
print(xtable(dat.lm,digits=1),booktabs=TRUE)
```

- Many more options, see `?xtable` and `?print.xtable`

# apsrtable

- Read the documentation - there are many options.

```{r}
require(apsrtable)
dat.lm2 <- lm(Y~X+0,data=dat)
apsrtable(dat.lm,dat.lm2)
```

# apsrtable

```{r fig.cap='', fig.width=6, fig.height=6}
library(png)
library(grid)
img <- readPNG("apsrtable.png")
grid.raster(img)
```

# stargazer

```{r}
require(stargazer)
stargazer(dat.lm,dat.lm2)
```

# stargazer

```{r fig.cap='', fig.width=6, fig.height=6}
img <- readPNG("stargazer.png")
grid.raster(img)
```

# Both

- Both packages are good (and can be supplemented with `xtable` when it is easier)

- Get pretty close to what you want with these packages, and then tweak the $\LaTeX$ directly.

# Plotting
- It's all about coordinate pairs.
- `plot(x,y)` plots the pairs of points in `x` and `y`
- Notable options:
    - `type` - determines whether you plot points, lines or whatnot
    - `pch` - determines plotting character
    - `xlim` - x limits of the plot (likewise for `y`)
    - `xlab` - label on the x-axis
    - `main` - main plot label
    - `col` - color
    - A massive number of options. Read the docs.
- Some objects respond specially to `plot`. Try `plot(dat.lm)`

# Tying it Together
```{r fig.cap='', fig.width=6, fig.height=6}
x <- seq(-1,1,.01)
y <- 3/4*(1-x^2)
plot(x,y, type="l", xlab="h", ylab="weight")
```

# Tying it Together
- $W$ is an $n\times p$ diagonal weighting matrix, $h$ is a "bandwidth".
- Diagonal entries are $\frac{3}{4}\cdot(1-d^2)\cdot 1_{\{|d|\le 1\}}$ where $d = \frac{X-c}{h}$
- $\hat{\beta}_c = (X'WX)^{-1}X'WY$
- Covariance matrix is $s^2(X'WX)^{-1}$

. . .

```{r}
loc.lin<-function(Y,X,c=0,bw=sd(X)/2){
  d<-(X-c)/bw
  W<-3/4*(1-d^2)*(abs(d)<1)
  W<-diag(W)
  X<-cbind(1,d)
  b<-solve(t(X)%*%W%*%X)%*%t(X)%*%W%*%Y
  sigma<-t(Y-X%*%b)%*%W%*%(Y-X%*%b)/(sum(diag(W)>0)-2)
  sigma<-solve(t(X)%*%W%*%X)*c(sigma)
  return(c(est=b[1],se=sqrt(diag(sigma))[1]))
}
```

# Fit the Surface

```{r fig.cap='', fig.width=6, fig.height=6}
X.est<-seq(0,5,.1)
dat.llm<-sapply(X.est, function(x) loc.lin(Y,X,c=x,bw=.25))
plot(X,Y,xlim=c(0,5),ylim=c(-50,50),pch=20)
lines(X.est,dat.llm[1,],col="red")
lines(X.est,dat.llm[1,]+1.96*dat.llm[2,],col="pink")
lines(X.est,dat.llm[1,]-1.96*dat.llm[2,],col="pink")
```
