% Regression Discontinuity
% Drew Dimmery <drewd@nyu.edu>
% April 4, 2014

# Structure
- RDD interpretation
- RDD estimation
- Placebo tests
- Sorting
- Other stuff

# Interpretation
- It's a LATE!
- A different kind of LATE!
- It can be interpreted as a weighted average over all units (Lee & Lemieux
  2010)
- $(W,U)$ are observed and unobserved factors which explain all heterogeneity.
- $X=c$ is the cutpoint on the running variable, $Y$ is the outcome
  $\lim_{\epsilon \downarrow 0} E[Y|X=c+\epsilon] - \lim_{\epsilon\uparrow 0}
  E[Y|X=c+\epsilon]$  
  $= \sum_{w,u} \tau(w,u) p(W=w,U=u|X=c)$  
  $= \sum_{w,u} \tau(w,u) {f(c|W=w,U=u) \over f(c)} p(W=w,U=u)$
- What does this mean?
- It's a weight of individual treatment effects weighted by the likelihood that
  a unit will lie near the threshhold on the running variable.
- Keep this in mind as you interpret results.

# Estimation
- If only someone wrote a package to do this...
- [http://github.com/ddimmery/rdd](http://github.com/ddimmery/rdd)
- The current best pracices is to use local polynomial regression.
- Typically linear
- There are also some interesting methods using randomization inference,
  though. (Cattaneo et al n.d.)

# Replication
- I'll be replicating the recent Meyersson paper that's been making noise.
- [Replication
  materials](http://www.econometricsociety.org/suppmat.asp?id=528&vid=82&iid=1&aid=7)
- The paper shows a (local) result that when Islamic parties won elections in
  Turkey, this resulted in better outcomes for women.
- Running variable: vote margin (but not exclusively 2 party system as in Lee)
- Outcome that we'll look at: high school education

. . .

```{r 7-data-setup}
require(foreign,quietly=TRUE)
d <- read.dta("regdata0.dta")
summary(d$iwm94)
```

# Explore data
- Plot the raw data.

. . .

```{r 7-raw-plot,fig.width=5,fig.height=5,fig.cap=''}
with(d,plot(iwm94,hischshr1520f,pch=19,cex=.2,xlim=c(-.5,.5)))
left.lm<-lm(hischshr1520f~iwm94,d,subset=iwm94<0)
right.lm<-lm(hischshr1520f~iwm94,d,subset=iwm94>=0)
left.x <- seq(-.5,0,.01)
right.x <- -left.x
lines(left.x,predict(left.lm,newd=data.frame(iwm94=left.x)),col="red")
lines(right.x,predict(right.lm,newd=data.frame(iwm94=right.x)),col="red")
```

# Estimation
- So the basic estimation would just take the difference of the intercepts from
  `left.lm` and `right.lm`.
- And there's an equivalency to just running a single regression as Cyrus
  showed in class.
- But I'm just going to use `rdd`

. . .

```{r 7-estimation}
require(rdd,quietly=TRUE)
rd.out<- RDestimate(hischshr1520f~iwm94,d)
rd.out
```

# Full Results
```{r 7-full-results}
summary(rd.out)
```

# Plot it
```{r 7-plot-it,fig.cap=''}
plot(rd.out,range=c(-.4,.4))
title(xlab="Islamic Party Vote Margin",ylab="Female High School Education Share")
```

# Placebo tests
- Do placebo tests on other covariates and other outcomes.
- They're "placebo" because there "shouldn't" be an effect on them (except
  occasionally by chance)

. . .

```{r 7-placebos}
# Age 19+
RDestimate(ageshr19~iwm94,d)[c("est","se")]
# Log Population
RDestimate(lpop1994~iwm94,d)[c("est","se")]
# Household Size
RDestimate(shhs~iwm94,d)[c("est","se")]
```

# More Placebos

```{r 7-more-placebos}
# Men in 2000
RDestimate(hischshr1520m~iwm94,d)[c("est","se")]
# Women in 1990 (pre-treatment)
RDestimate(c90hischshr1520f~iwm94,d)[c("est","se")]
# Men in 1990 (pre-treatment)
RDestimate(c90hischshr1520m~iwm94,d)[c("est","se")]
```

# Sorting
- As Cyrus discussed, density tests are also a good way to examine the
  possibility of sorting.

. . .

```{r 7-dcd}
DCdensity(d$iwm94,verbose=TRUE,plot=FALSE)
```

# Density Plot

```{r 7-dcd-plot,fig.cap=''}
DCdensity(d$iwm94)
```

# Fuzzy designs
- I don't have an example for this, but it's quite easy.
- Do it the same way as before, but with `RDestimate(Y~runvar+treatment)`

# Overall
- Some big things for RDD:
    - Lots of plots
    - Think about locality in interpretation
    - Use your covariates for robustness/placebo tests
    - Everything should be robust to different bandwidths, etc
    - If effects start disappearing as bw goes down, that's a bad sign.
    - Your bandwidth is probably to wide.
- If there's still more time, maybe I'll go through some high points of the
  `rdd` code.
